\morus: Biases in the Stream

### 2018-05-10 morusAC_01_Introduction.tex
minor consistency fix: MORUS

\morus is one of the three finalists for use-case (2), together with \cipher{OCB} and \cipher{AEGIS}.
This family of authenticated ciphers by Wu and Huang \cite{MORUS} provides three main variants:
\morus[640] with a 128-bit key and \morus[1280] with either a 128-bit or a 256-bit key.
The design approach is reminiscent of classical stream cipher designs and
continuously updates a relatively large state with a few fast operations.
\morus can be efficiently implemented in both software and hardware;
in particular, the designers claim that the software performance even surpasses
\cipher{AES-GCM} implementations using Intel's \cipher{AES-NI} instructions,
and that \morus is the fastest authenticated cipher not using \cipher{AES-NI} \cite{MORUS}.

\subsubsection*{Related Work.}
In the \morus submission document, the designers discuss the security of \morus against
several attacks, including algebraic, differential, and guess-and-determine attacks.
The main focus is on differential properties, and not many details are given for other attack vectors.
In third-party analysis,
Mileva et~al.~\cite{balkancryptsecMilevaDV15} propose a distinguisher in the nonce-reuse setting and practically evaluate the differential behaviour of toy variants of \morus.
Dwivedi et~al.~\cite{cryptoeprint:2016:1053} discuss the applicability of SAT solvers for state recovery, but the resulting complexity of $2^{370}$ for \morus[640] is well beyond the security claim.
Dwivedi et~al.~\cite{secryptDwivediMW17} also propose key-recovery attacks for \morus[1280] if initialization is reduced to 3.6 out of 16 rounds, and discuss the security of \morus against internal differentials and rotational cryptanalysis.
Salam et~al.~\cite{trustcomSalamSBDPW17} apply cube attacks to obtain distinguishers for up to 5 out of 16 rounds of the initialization of \morus[1280] with negligible complexity.
Additionally, 
Kales et~al.~\cite{cryptoeprint:2017:1137} and
Vaudenay and Viz\'{a}r \cite{cryptoeprint:2017:1147}
independently propose state-recovery and forgery attacks on \morus in a nonce-misuse setting with negligible data and time complexities.
Finally, a keystream bias of a similar nature as our main attack was uncovered by Minaud \cite{sacryptMinaud14} on the authenticated cipher \cipher{AEGIS} \cite{AEGIS,sacryptWuP13}, another CAESAR finalist which shares the same overall structure as \morus, but uses a very different, \cipher{AES}-based state update function.

\subsubsection*{Our Contributions.}
We analyze the security of \morus and present a key\-stream distinguisher based on
linear approximations for full \morus[1280].
In addition, we provide results for round-reduced \morus with reduced initialization or finalization.

Our main result is a linear approximation \cite{eurocryptMatsui93,eurocryptMatsuiY92} linking ciphertext bits of five consecutive encryption blocks.
%As a consequence of this trail, we exhibit a biased linear combination of ciphertext bits, which holds across any given sequence of five consecutive encryption steps.
%Technically, this assumes that the inner state of the cipher at the start of the trail is uniformly random, but this is a fair assumption since the inner state should be indistinguishable from uniform randomness by design.
%Moreover, that is the only assumption required; in particular, the secret key or the nonce used to initialize the inner state of MORUS are irrelevant: the bias is independent of their value.
More precisely, we show that when encrypting a fixed plaintext block multiple times, a linear correlation exists between some output bits of the cipher.
Moreover, the bias depends purely on the plaintext, and not on the secret key of the cipher.
In principle, this property could be used to recover unknown bits of a plaintext encrypted a large number of times, provided an initial segment of the plaintext is known.
For \morus[1280] with 256-bit keys, the linear bias is $2^{-73}$
\todo{double-check bias vs. correlation everywhere}
and can be exploited using about $2^{152}$ encrypted blocks.
\todo{double-check}
We note that our attack is nonce-respecting, and that rekeying does not prevent it: the biases are independent of the secret encryption key and nonce, and can be exploited as long as a given plaintext block is encrypted sufficiently often, regardless of whether each encryption uses a different key.

To obtain this result, we propose a simplified abstraction of \morus, called \minimorus, which takes advantage of certain rotational symmetries in \morus.
We observe that the actual bias of the approximation is better than suggested by the construction of the characteristics.
To confirm the validity of our analysis,
we practically verified the bias of the linear approximation for \minimorus and of partial approximations for \morus.
In addition, we symbolically evaluated the relevant equations.

In addition to this attack on full \morus, we also analyze the security of \morus with round-reduced initialization (in a nonce-misuse setting) or finalization.
While this analysis does not threaten full \morus, it complements the main result to provide a better understanding of the \morus design approach.
We present a forgery attack for round-reduced \morus[1280] with success probability $2^{-88}$ for a 128-bit tag if the finalization is reduced to 3 out of 10 rounds. This nonce-respecting attack is based on a differential analysis of the padding rule.
The second result targets a round-reduced initialization with 10 out of 16 rounds, and extends a state-recovery attack (e.g., in a nonce-misuse setting) into a key-recovery attack.

\subsubsection*{Outline.}
This paper is organized as follows.
We first provide a brief description of \morus in \Cref{sec/Preliminaries}.
In \Cref{sec/introminimorus}, we introduce \minimorus, an abstraction of \morus based on a certain class of rotational symmetries.
We analyze this simplified scheme in \Cref{sec/minitrails} and provide a ciphertext-only linear approximation with a weight of 16.
We then extend our result to the full scheme in \Cref{sec/fulltrails}, showing a bias in the keystream over 5 rounds,
and discuss the implications of our observation for the security of \morus in \Cref{sec/Discussion}.
In \Cref{sec/IniFin}, we present our results on the security of \morus with round-reduced initialization (in a nonce-misuse setting) or finalization.
Finally, we conclude in \Cref{sec/Conclusion}.

>>>>

\MORUS is one of the three finalists for use-case (2), together with \cipher{OCB} and \cipher{AEGIS}.
This family of authenticated ciphers by Wu and Huang \cite{MORUS} provides three main variants:
\MORUS[640] with a 128-bit key and \MORUS[1280] with either a 128-bit or a 256-bit key.
The design approach is reminiscent of classical stream cipher designs and
continuously updates a relatively large state with a few fast operations.
\MORUS can be efficiently implemented in both software and hardware;
in particular, the designers claim that the software performance even surpasses
\cipher{AES-GCM} implementations using Intel's \cipher{AES-NI} instructions,
and that \MORUS is the fastest authenticated cipher not using \cipher{AES-NI} \cite{MORUS}.

\subsubsection*{Related Work.}
In the \MORUS submission document, the designers discuss the security of \MORUS against
several attacks, including algebraic, differential, and guess-and-determine attacks.
The main focus is on differential properties, and not many details are given for other attack vectors.
In third-party analysis,
Mileva et~al.~\cite{balkancryptsecMilevaDV15} propose a distinguisher in the nonce-reuse setting and practically evaluate the differential behaviour of toy variants of \MORUS.
Dwivedi et~al.~\cite{cryptoeprint:2016:1053} discuss the applicability of SAT solvers for state recovery, but the resulting complexity of $2^{370}$ for \MORUS[640] is well beyond the security claim.
Dwivedi et~al.~\cite{secryptDwivediMW17} also propose key-recovery attacks for \MORUS[1280] if initialization is reduced to 3.6 out of 16 rounds, and discuss the security of \MORUS against internal differentials and rotational cryptanalysis.
Salam et~al.~\cite{trustcomSalamSBDPW17} apply cube attacks to obtain distinguishers for up to 5 out of 16 rounds of the initialization of \MORUS[1280] with negligible complexity.
Additionally, 
Kales et~al.~\cite{cryptoeprint:2017:1137} and
Vaudenay and Viz\'{a}r \cite{cryptoeprint:2017:1147}
independently propose state-recovery and forgery attacks on \MORUS in a nonce-misuse setting with negligible data and time complexities.
Finally, a keystream bias of a similar nature as our main attack was uncovered by Minaud \cite{sacryptMinaud14} on the authenticated cipher \cipher{AEGIS} \cite{AEGIS,sacryptWuP13}, another CAESAR finalist which shares the same overall structure as \MORUS, but uses a very different, \cipher{AES}-based state update function.

\subsubsection*{Our Contributions.}
We analyze the security of \MORUS and present a key\-stream distinguisher based on
linear approximations for full \MORUS[1280].
In addition, we provide results for round-reduced \MORUS with reduced initialization or finalization.

Our main result is a linear approximation \cite{eurocryptMatsui93,eurocryptMatsuiY92} linking ciphertext bits of five consecutive encryption blocks.
%As a consequence of this trail, we exhibit a biased linear combination of ciphertext bits, which holds across any given sequence of five consecutive encryption steps.
%Technically, this assumes that the inner state of the cipher at the start of the trail is uniformly random, but this is a fair assumption since the inner state should be indistinguishable from uniform randomness by design.
%Moreover, that is the only assumption required; in particular, the secret key or the nonce used to initialize the inner state of MORUS are irrelevant: the bias is independent of their value.
More precisely, we show that when encrypting a fixed plaintext block multiple times, a linear correlation exists between some output bits of the cipher.
Moreover, the bias depends purely on the plaintext, and not on the secret key of the cipher.
In principle, this property could be used to recover unknown bits of a plaintext encrypted a large number of times, provided an initial segment of the plaintext is known.
For \MORUS[1280] with 256-bit keys, the linear bias is $2^{-73}$
\todo{double-check bias vs. correlation everywhere}
and can be exploited using about $2^{152}$ encrypted blocks.
\todo{double-check}
We note that our attack is nonce-respecting, and that rekeying does not prevent it: the biases are independent of the secret encryption key and nonce, and can be exploited as long as a given plaintext block is encrypted sufficiently often, regardless of whether each encryption uses a different key.

To obtain this result, we propose a simplified abstraction of \MORUS, called \MiniMORUS, which takes advantage of certain rotational symmetries in \MORUS.
We observe that the actual bias of the approximation is better than suggested by the construction of the characteristics.
To confirm the validity of our analysis,
we practically verified the bias of the linear approximation for \MiniMORUS and of partial approximations for \MORUS.
In addition, we symbolically evaluated the relevant equations.

In addition to this attack on full \MORUS, we also analyze the security of \MORUS with round-reduced initialization (in a nonce-misuse setting) or finalization.
While this analysis does not threaten full \MORUS, it complements the main result to provide a better understanding of the \MORUS design approach.
We present a forgery attack for round-reduced \MORUS[1280] with success probability $2^{-88}$ for a 128-bit tag if the finalization is reduced to 3 out of 10 rounds. This nonce-respecting attack is based on a differential analysis of the padding rule.
The second result targets a round-reduced initialization with 10 out of 16 rounds, and extends a state-recovery attack (e.g., in a nonce-misuse setting) into a key-recovery attack.

\subsubsection*{Outline.}
This paper is organized as follows.
We first provide a brief description of \MORUS in \Cref{sec/Preliminaries}.
In \Cref{sec/introminimorus}, we introduce \MiniMORUS, an abstraction of \MORUS based on a certain class of rotational symmetries.
We analyze this simplified scheme in \Cref{sec/minitrails} and provide a ciphertext-only linear approximation with a weight of 16.
We then extend our result to the full scheme in \Cref{sec/fulltrails}, showing a bias in the keystream over 5 rounds,
and discuss the implications of our observation for the security of \MORUS in \Cref{sec/Discussion}.
In \Cref{sec/IniFin}, we present our results on the security of \MORUS with round-reduced initialization (in a nonce-misuse setting) or finalization.
Finally, we conclude in \Cref{sec/Conclusion}.

### 2018-05-10 morusAC_02_Preliminaries.tex
minor consistency fix: MORUS

MORUS \cite{MORUS} is a family of authenticated ciphers with two internal state sizes: 640 and 1280 bits.
Each member is called \cipher{MORUS640} and \cipher{MORUS1280}, respectively.

An instance of MORUS is parametrized by a secret key. During encryption, it takes as input a plaintext message and possibly some associated data, and outputs a ciphertext together with an authentication tag.
In the recommended parameters, \cipher{MORUS640} supports 128-bit keys and \cipher{MORUS1280} supports both 128-bit and 256-bit keys. The tag size is 128 bits or shorter. The designers strongly recommend using a 128-bit tag.

\subsection{Specification of MORUS}
\label{subsec/Spec}

The internal state of MORUS is composed of five registers. Let $q$ be the register size, where $q = 128$ for \cipher{MORUS640} and $q=256$ for \cipher{MORUS1280}. Let $S_i, i \in\{0,1,2,3,4\}$ be a register. Then the internal state of MORUS may be represented as $S_0\|S_1\|S_2\|S_3\|S_4$.
Registers are themselves divided into four $q/4$-bit \emph{words}.
Throughout the paper, we denote the word size by $w = q/4$, i.e. $w=32$ for \cipher{MORUS640} and $w=64$ for \cipher{MORUS1280}.

The encryption process of MORUS consists of four parts: initialization, associated data processing, encryption, and finalization. 
During the initialization phase, the value of the state is initialized using a key and nonce.
Associated data block and plaintext blocks are then processed block by block.
Then the internal state undergoes the finalization phase, which outputs the authentication tag.

Every part of this process relies on iterating the \StateUpdate{} function at the core of MORUS. Each call to the \StateUpdate{} function is called a step.
In order to denote the internal state of the cipher at the $t$-th step, we use a superscript to denote the step number: the internal state at step $t$ is denoted by $S^t_0\|S^t_1\|S^t_2\|S^t_3\|S^t_4$, where $t$ is $-16$ before the initialization and $t=0$ after the initialization.

@@

\StateUpdate{} is composed of 5 rounds with similar operations. The additional input $m^t$ is used in rounds 2 to 5, but not in round 1. Each round uses the word-wise left circular shift operation, which divides a $q$-bit register value into 4 words of $w = q/4$ bits, and performs a left circular shift operation on each $w$-bit word. The word-wise circular shift constants $b_i$ in round $i$ are defined in \Cref{Tbl:rcon}. Left circular shifts on a whole $q$-bit register are also used. The register-wise circular shift constants are also listed in \Cref{Tbl:rcon}.
\begin{table}[!htb]
\centering
\caption{Left Circular Shift Constants of MORUS} \label{Tbl:rcon}
\begin{tabular}{r||ccccc|ccccc} \hline
& \multicolumn{5}{c|}{Word-wise circular shift} & \multicolumn{5}{c}{Register-wise circular shift} \\
                  & \makebox[2em]{$b_0$} & \makebox[2em]{$b_1$} & \makebox[2em]{$b_2$} & \makebox[2em]{$b_3$} & \makebox[2em]{$b_4$} & \makebox[2em]{$b'_0$} & \makebox[2em]{$b'_1$} & \makebox[2em]{$b'_2$} & \makebox[2em]{$b'_3$} & \makebox[2em]{$b'_4$} \\ \hline
\cipher{MORUS640} &  5 & 31 &  7 & 22 & 13 & 32 &  64 &  96 &  64 & 32 \\
\cipher{MORUS1280}& 13 & 46 & 38 &  7 &  4 & 64 & 128 & 192 & 128 & 64 \\ \hline
\end{tabular}
\end{table}

We denote those register-wise and word-wise left circular shift by $x \lll y$ and $x \lll_W y$, respectively. (Those are denoted by \texttt{<<<} and \texttt{Rotl\_xxx\_yy} respectively in the \cipher{MORUS} design document.)

@@

\subsubsection{Initialization.}
The initialization of \cipher{MORUS640} consists of loading a 128-bit key $K_{128}$ and a 128-bit nonce $IV_{128}$ into the state, and running \StateUpdate{} for 16 steps. The key and nonce are loaded into the state as follows:
\begin{align*}
S^{-16}_0 = \textrm{IV}_{128}, &&
S^{-16}_1 = K_{128}, &&
S^{-16}_2 = 1^{128}, &&
S^{-16}_3 = const_0, &&
S^{-16}_4 = const_1.
\end{align*}
The internal state is then updated 16 times with  $\StateUpdate{}(S^t,0)$ for $t=-16,-15,\ldots,-1$. Finally, the key is XORed into the state again with $S^0_1 \leftarrow S^0_1 \oplus K_{128}$.

The initialization of \cipher{MORUS1280} is slightly different from that of \cipher{MORUS640} due to the difference in register size and the two possible key sizes:
\begin{itemize}
  \item[-] when the key size is 128 bits, $K = K_{128} \|K_{128}$.
  \item[-] when the key size is 256 bits, $K = K_{256}$.
\end{itemize}
The state is initialized as:
\begin{align*}
S^{-16}_0 &= \textrm{IV}_{128} \| 0^{128}, &
S^{-16}_1 &= K, &
S^{-16}_2 = 1^{256}, \\
S^{-16}_3 &= 0^{256}, &
S^{-16}_4 &= const_0\|const_1.
\end{align*}
After updating the state 16 times with \StateUpdate{}, the state is updated with $S^0_1 \leftarrow S^0_1 \oplus K$.

\subsubsection{Associated Data Processing.}
After initialization, the associated data $A$ is processed. For the padding, if the last associated data block is not a full block,  it is padded to 128 bits and 256 bits for \cipher{MORUS640} and \cipher{MORUS1280} respectively, using 0 bits. The padded full block is then used to update the state. If the length of $A$, denoted by `${\it adlen}$' is 0, the associated data processing phase is skipped.

The state is then updated with $S^{t+1} \leftarrow \StateUpdate{}(S^t, A^t)$ for $t=0,1,\ldots,\ell$ where $\ell = \lceil \frac{{\it adlen}}{128} \rceil -1$ and $\ell = \lceil \frac{{\it adlen}}{256} \rceil -1$ for \cipher{MORUS640} and \cipher{MORUS1280}, respectively.

\subsubsection{Encryption.}
At each encryption step, a 16-byte (resp. 32-byte) message block $M_t$ is used to update the state in \cipher{MORUS640} (resp. \cipher{MORUS1280}), and a ciphertext $C_t$ is produced.

If the last message block is not a full block, 0 bits are used to pad it to 128 and 256 bits for \cipher{MORUS640} and \cipher{MORUS1280}, respectively, and the padded full block is used to update the state. But only the partial block is encrypted. Note that if the message length denoted by `${\it msglen}$' is 0, encryption is skipped.

In \cipher{MORUS640}, Let $u = \lceil \frac{{\it adlen}}{128} \rceil$ and $v = \lceil \frac{{\it msglen}}{128} \rceil$. The following is performed for $t=0, 1, \ldots, v-1$:
\begin{align*}
C^t &\leftarrow M^t \oplus S^{u+t}_0 \oplus (S^{u+t}_1 \lll 96) \oplus (S^{u+t}_2 \cdot S^{u+t}_3),\\
S^{u+t+1} &\leftarrow \StateUpdate{}(S^{u+t},M^t).
\end{align*}

For \cipher{MORUS1280}, $u = \lceil \frac{{\it adlen}}{256} \rceil$ and $v = \lceil \frac{{\it msglen}}{256} \rceil$. The ciphertext block $C^t$ is computed as $C^t \leftarrow M^t \oplus S^{u+t}_0 \oplus (S^{u+t}_1 \lll 192) \oplus (S^{u+t}_2 \cdot S^{u+t}_3)$.

\subsubsection{Finalization.}
The finalization phase generates the authentication tag $T$ using 10 more \StateUpdate{} steps. We only discuss the case where $T$ is not truncated. The associated data length and the message length are used to update the state as follows.
\begin{enumerate}
\item ${\it tmp} \leftarrow {\it adlen}\|{\it msglen}$ for \cipher{MORUS640} and ${\it tmp} \leftarrow {\it adlen}\|{\it msglen}\|0^{128}$ for \cipher{MORUS1280}, where ${\it adlen}$ and ${\it msglen}$ are represented as 64-bit integers.
\item $S^{u+v}_4 \leftarrow S^{u+v}_4 \oplus S^{u+v}_0.$
\item For $t = u+v, u+v+1, \ldots, u+v+9,$ compute
%\begin{equation*}
$S^{t+1} \leftarrow \StateUpdate{} (S^t, {\it tmp})$.
%\end{equation*}
\item For \cipher{MORUS640}, $T = S^{u+v+10}_0 \oplus (S^{u+v+10}_1 \lll 96) \oplus ( S^{u+v+10}_2 \cdot S^{u+v+10}_3)$. For \cipher{MORUS1280}, $S^{u+v+10}_0 \oplus (S^{u+v+10}_1 \lll 192) \oplus ( S^{u+v+10}_2 \cdot S^{u+v+10}_3)$ is computed and the least significant 128 bits are produced as a tag $T$.
\end{enumerate}

\subsubsection{Security.}
With a 128-bit tag, integrity is claimed up to 128 bits and confidentiality is claimed up to the number of bits of the key.
\begin{table}[!htb]
\centering
\caption{Security Goals of MORUS.}
\begin{tabular}{ccc}\hline
                        & Confidentiality (bits) & Integrity (bits) \\ \hline
\cipher{MORUS640}-128   & 128                    & 128              \\
\cipher{MORUS1280}-128  & 128                    & 128              \\
\cipher{MORUS1280}-256  & 256                    & 128              \\ \hline
\end{tabular}
\end{table}

@@

We also recall the following notation from the previous section, where an \emph{encryption step} refers to one call to the \StateUpdate{} function:\\
\centerline{
\def\arraystretch{1.25}
\begin{tabular}{lcl}
$C^t$ &:& the ciphertext block output during the $t$-th encryption step.\\
$C^t_j$ &:& the $j$-th bit of $C^t$, with $C^t_0$ being the rightmost bit.\\
$S^t_i$ &:& the $i$-th register at the beginning of $t$-th encryption step.\\
$S^t_{i,j}$ &:& the $j$-th bit of $S^t_i$, with $S^t_{i,0}$ being the rightmost bit.
\end{tabular}
}
In the above notation, bit positions are always taken modulo the register size, i.e. 128 for \cipher{MORUS640} and 256 for \cipher{MORUS1280}.

@@

%Finally, let $x$ denote a register. Recall that registers of \cipher{MORUS} are composed of four words. We now define register-wise and word-wise rotations (denoted by \texttt{<<<} and \texttt{Rotl\_xxx\_yy} respectively in the \cipher{MORUS} design document):\\
%\centerline{
%\def\arraystretch{1.25}
%\begin{tabular}{lcl}
%$x \rotl y$ &:& circular shift of $x$ by $y$ bits to the left.\\
%$x \rotlxy y$ &:& circular shift of each word within $x$ by $y$ bits to the left.\\
%\end{tabular}
%}
%
%The $b_i$'s denote rotation constants, whose value depends on which version of \cipher{MORUS} is considered. Throughout, $w$ denotes the  word size, i.e. $w=32$ for \cipher{MORUS640} and $w=64$ for \cipher{MORUS1280}.

>>>>

\MORUS \cite{MORUS} is a family of authenticated ciphers with two internal state sizes: 640 and 1280 bits.
Each member is called \MORUS[640] and \MORUS[1280], respectively.

An instance of \MORUS is parametrized by a secret key. During encryption, it takes as input a plaintext message and possibly some associated data, and outputs a ciphertext together with an authentication tag.
In the recommended parameters, \MORUS[640] supports 128-bit keys and \MORUS[1280] supports both 128-bit and 256-bit keys. The tag size is 128 bits or shorter. The designers strongly recommend using a 128-bit tag.

\subsection{Specification of \MORUS}
\label{subsec/Spec}

The internal state of \MORUS is composed of five registers. Let $q$ be the register size, where $q = 128$ for \MORUS[640] and $q=256$ for \MORUS[1280]. Let $S_i, i \in\{0,1,2,3,4\}$ be a register. Then the internal state of \MORUS may be represented as $S_0\|S_1\|S_2\|S_3\|S_4$.
Registers are themselves divided into four $q/4$-bit \emph{words}.
Throughout the paper, we denote the word size by $w = q/4$, i.e. $w=32$ for \MORUS[640] and $w=64$ for \MORUS[1280].

The encryption process of \MORUS consists of four parts: initialization, associated data processing, encryption, and finalization. 
During the initialization phase, the value of the state is initialized using a key and nonce.
Associated data block and plaintext blocks are then processed block by block.
Then the internal state undergoes the finalization phase, which outputs the authentication tag.

Every part of this process relies on iterating the \StateUpdate{} function at the core of \MORUS. Each call to the \StateUpdate{} function is called a step.
In order to denote the internal state of the cipher at the $t$-th step, we use a superscript to denote the step number: the internal state at step $t$ is denoted by $S^t_0\|S^t_1\|S^t_2\|S^t_3\|S^t_4$, where $t$ is $-16$ before the initialization and $t=0$ after the initialization.

@@

\StateUpdate{} is composed of 5 rounds with similar operations. The additional input $m^t$ is used in rounds 2 to 5, but not in round 1. Each round uses the word-wise left circular shift operation, which divides a $q$-bit register value into 4 words of $w = q/4$ bits, and performs a left circular shift operation on each $w$-bit word. The word-wise circular shift constants $b_i$ in round $i$ are defined in \Cref{Tbl:rcon}. Left circular shifts on a whole $q$-bit register are also used. The register-wise circular shift constants are also listed in \Cref{Tbl:rcon}.
\begin{table}[!htb]
\centering
\caption{Left Circular Shift Constants of \texorpdfstring{\MORUS}{MORUS}} \label{Tbl:rcon}
\begin{tabular}{r||ccccc|ccccc} \hline
& \multicolumn{5}{c|}{Word-wise circular shift} & \multicolumn{5}{c}{Register-wise circular shift} \\
                  & \makebox[2em]{$b_0$} & \makebox[2em]{$b_1$} & \makebox[2em]{$b_2$} & \makebox[2em]{$b_3$} & \makebox[2em]{$b_4$} & \makebox[2em]{$b'_0$} & \makebox[2em]{$b'_1$} & \makebox[2em]{$b'_2$} & \makebox[2em]{$b'_3$} & \makebox[2em]{$b'_4$} \\ \hline
\MORUS[640]  &  5 & 31 &  7 & 22 & 13 & 32 &  64 &  96 &  64 & 32 \\
\MORUS[1280] & 13 & 46 & 38 &  7 &  4 & 64 & 128 & 192 & 128 & 64 \\ \hline
\end{tabular}
\end{table}

We denote those register-wise and word-wise left circular shift by $x \lll y$ and $x \lll_W y$, respectively. (Those are denoted by \texttt{<<<} and \texttt{Rotl\_xxx\_yy} respectively in the \MORUS design document.)

@@

\subsubsection{Initialization.}
The initialization of \MORUS[640] consists of loading a 128-bit key $K_{128}$ and a 128-bit nonce $IV_{128}$ into the state, and running \StateUpdate{} for 16 steps. The key and nonce are loaded into the state as follows:
\begin{align*}
S^{-16}_0 = \textrm{IV}_{128}, &&
S^{-16}_1 = K_{128}, &&
S^{-16}_2 = 1^{128}, &&
S^{-16}_3 = const_0, &&
S^{-16}_4 = const_1.
\end{align*}
The internal state is then updated 16 times with  $\StateUpdate{}(S^t,0)$ for $t=-16,-15,\ldots,-1$. Finally, the key is XORed into the state again with $S^0_1 \leftarrow S^0_1 \oplus K_{128}$.

The initialization of \MORUS[1280] is slightly different from that of \MORUS[640] due to the difference in register size and the two possible key sizes:
\begin{itemize}
  \item[-] when the key size is 128 bits, $K = K_{128} \|K_{128}$.
  \item[-] when the key size is 256 bits, $K = K_{256}$.
\end{itemize}
The state is initialized as:
\begin{align*}
S^{-16}_0 &= \textrm{IV}_{128} \| 0^{128}, &
S^{-16}_1 &= K, &
S^{-16}_2 = 1^{256}, \\
S^{-16}_3 &= 0^{256}, &
S^{-16}_4 &= const_0\|const_1.
\end{align*}
After updating the state 16 times with \StateUpdate{}, the state is updated with $S^0_1 \leftarrow S^0_1 \oplus K$.

\subsubsection{Associated Data Processing.}
After initialization, the associated data $A$ is processed. For the padding, if the last associated data block is not a full block,  it is padded to 128 bits and 256 bits for \MORUS[640] and \MORUS[1280] respectively, using 0 bits. The padded full block is then used to update the state. If the length of $A$, denoted by `${\it adlen}$' is 0, the associated data processing phase is skipped.

The state is then updated with $S^{t+1} \leftarrow \StateUpdate{}(S^t, A^t)$ for $t=0,1,\ldots,\ell$ where $\ell = \lceil \frac{{\it adlen}}{128} \rceil -1$ and $\ell = \lceil \frac{{\it adlen}}{256} \rceil -1$ for \MORUS[640] and \MORUS[1280], respectively.

\subsubsection{Encryption.}
At each encryption step, a 16-byte (resp. 32-byte) message block $M_t$ is used to update the state in \MORUS[640] (resp. \MORUS[1280]), and a ciphertext $C_t$ is produced.

If the last message block is not a full block, 0 bits are used to pad it to 128 and 256 bits for \MORUS[640] and \MORUS[1280], respectively, and the padded full block is used to update the state. But only the partial block is encrypted. Note that if the message length denoted by `${\it msglen}$' is 0, encryption is skipped.

In \MORUS[640], Let $u = \lceil \frac{{\it adlen}}{128} \rceil$ and $v = \lceil \frac{{\it msglen}}{128} \rceil$. The following is performed for $t=0, 1, \ldots, v-1$:
\begin{align*}
C^t &\leftarrow M^t \oplus S^{u+t}_0 \oplus (S^{u+t}_1 \lll 96) \oplus (S^{u+t}_2 \cdot S^{u+t}_3),\\
S^{u+t+1} &\leftarrow \StateUpdate{}(S^{u+t},M^t).
\end{align*}

For \MORUS[1280], $u = \lceil \frac{{\it adlen}}{256} \rceil$ and $v = \lceil \frac{{\it msglen}}{256} \rceil$. The ciphertext block $C^t$ is computed as $C^t \leftarrow M^t \oplus S^{u+t}_0 \oplus (S^{u+t}_1 \lll 192) \oplus (S^{u+t}_2 \cdot S^{u+t}_3)$.

\subsubsection{Finalization.}
The finalization phase generates the authentication tag $T$ using 10 more \StateUpdate{} steps. We only discuss the case where $T$ is not truncated. The associated data length and the message length are used to update the state as follows.
\begin{enumerate}
\item ${\it tmp} \leftarrow {\it adlen}\|{\it msglen}$ for \MORUS[640] and ${\it tmp} \leftarrow {\it adlen}\|{\it msglen}\|0^{128}$ for \MORUS[1280], where ${\it adlen}$ and ${\it msglen}$ are represented as 64-bit integers.
\item $S^{u+v}_4 \leftarrow S^{u+v}_4 \oplus S^{u+v}_0.$
\item For $t = u+v, u+v+1, \ldots, u+v+9,$ compute
%\begin{equation*}
$S^{t+1} \leftarrow \StateUpdate{} (S^t, {\it tmp})$.
%\end{equation*}
\item For \MORUS[640], $T = S^{u+v+10}_0 \oplus (S^{u+v+10}_1 \lll 96) \oplus ( S^{u+v+10}_2 \cdot S^{u+v+10}_3)$. For \MORUS[1280], $S^{u+v+10}_0 \oplus (S^{u+v+10}_1 \lll 192) \oplus ( S^{u+v+10}_2 \cdot S^{u+v+10}_3)$ is computed and the least significant 128 bits are produced as a tag $T$.
\end{enumerate}

\subsubsection{Security.}
With a 128-bit tag, integrity is claimed up to 128 bits and confidentiality is claimed up to the number of bits of the key. See Table~\ref{Tbl/security} for summary.
\begin{table}[!htb]
\centering
\caption{Security goals of \MORUS.} \label{Tbl/security}
\begin{tabular}{ccc}\hline
                        & Confidentiality (bits) & Integrity (bits) \\ \hline
\MORUS[640-128]   & 128                    & 128              \\
\MORUS[1280-128]  & 128                    & 128              \\
\MORUS[1280-256]  & 256                    & 128              \\ \hline
\end{tabular}
\end{table}

@@

We also recall the following notation from the previous section, where an \emph{encryption step} refers to one call to the \StateUpdate{} function:\\
\centerline{
\def\arraystretch{1.25}
\begin{tabular}{lcl}
$C^t$ &:& the ciphertext block output during the $t$-th encryption step.\\
$C^t_j$ &:& the $j$-th bit of $C^t$, with $C^t_0$ being the rightmost bit.\\
$S^t_i$ &:& the $i$-th register at the beginning of $t$-th encryption step.\\
$S^t_{i,j}$ &:& the $j$-th bit of $S^t_i$, with $S^t_{i,0}$ being the rightmost bit.
\end{tabular}
}
In the above notation, bit positions are always taken modulo the register size, i.e. 128 for \MORUS[640] and 256 for \MORUS[1280].

@@

%Finally, let $x$ denote a register. Recall that registers of \MORUS are composed of four words. We now define register-wise and word-wise rotations (denoted by \texttt{<<<} and \texttt{Rotl\_xxx\_yy} respectively in the \MORUS design document):\\
%\centerline{
%\def\arraystretch{1.25}
%\begin{tabular}{lcl}
%$x \rotl y$ &:& circular shift of $x$ by $y$ bits to the left.\\
%$x \rotlxy y$ &:& circular shift of each word within $x$ by $y$ bits to the left.\\
%\end{tabular}
%}
%
%The $b_i$'s denote rotation constants, whose value depends on which version of \MORUS is considered. Throughout, $w$ denotes the  word size, i.e. $w=32$ for \MORUS[640] and $w=64$ for \MORUS[1280].
